---
title: "Development guide"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4
    toc: true
    toc_depth: 4
    number_sections: false
    df_print: paged # default, kable, tibble
    fig_caption: true
  pdf_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Development guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Here we describe some  general guidelines for maintenance and development of `gwloggeR`.

## General

`gwloggeR` is developed as part of a larger git-repository: https://github.com/DOV-Vlaanderen/groundwater-logger-validation.

So first we need to checkout this repository somewhere locally.

```{}
git checkout https://github.com/DOV-Vlaanderen/groundwater-logger-validation.git
```

### Projects

The repository consists of 3 project roots:

* [`./src/r`](https://github.com/DOV-Vlaanderen/groundwater-logger-validation/tree/master/src/r) which is mainly used for data analysis with R. The project file `groundwater.rproj`, when opened in Rstudio, will also load [`src/r/.Rprofile`](https://github.com/DOV-Vlaanderen/groundwater-logger-validation/blob/master/src/r/.Rprofile) which will setup the environment for data analysis.
* [`./gwloggeR`](https://github.com/DOV-Vlaanderen/groundwater-logger-validation/tree/master/gwloggeR) is the `gwloggeR` package source repository used for development and testing. Open `gwloggeR.Rproj` with Rstudio.
* [`./gwloggeR.data`](https://github.com/DOV-Vlaanderen/groundwater-logger-validation/tree/master/gwloggeR.data) is a small convenience package developed for uniform data loading from the `./data/raw` directory. Currently it has no explicit Rstudio project file.

Note that cross-linking of files between projects based on relative paths is allowed since all projects reside in the same repository.

Always double-check the `.Rprofile` file and adjust the paths accordingly.

## Testing

There is a dedicated [README](https://github.com/DOV-Vlaanderen/groundwater-logger-validation/tree/master/gwloggeR/tests) about testing `gwloggeR`.

Before releasing a version, make sure to not only run the `testthat` procedures (`Ctrl` + `Shift` + `T`), but also the extensive _analytics_ tests to make sure that the package results conform with past expected results (i.e. the package still operates as intended on previously analyzed data).

## Documentation

The `gwloggeR` documentation is built using `roxygen2`. If you are using RStudio and `roxygen2` is installed, then pressing `Ctrl` + `Shift` + `B` will:

* Update Rd files (i.e. R documentation files)
* Update NAMESPACE file
* Update vignettes

This command will also install `gwloggeR` in the R `lib\` directory and restart R.

On each update of `gwloggeR` source files, make sure to run this before committing.

### README.Rmd

This is a special case that must be knitted to `README.md`. In RStudio one can use `Ctrl` + `Shift` + `K`. Knitting must be done before the website is built.

## Website

For `gwloggeR` [website](https://dov-vlaanderen.github.io/groundwater-logger-validation/gwloggeR) we use `pkgdown`. 

We do not use [`travis`](https://pkgdown.r-lib.org/reference/deploy_site_github.html) for automatic website deployment. Each time one updates files that have an impact on the website, one should manually execute the following commands:

```{r eval=FALSE}
pkgdown::clean_site()
pkgdown::build_site()
```

On each update of `gwloggeR` source files, make sure to run these commands before committing.

## Deployment

Before deployment of a new version, certain steps need to be taken.

* Update the [Documentation].
* Make sure you complete [Testing], and also re-run the more elaborate _analytics_ tests.
* Update the [News.md](https://github.com/DOV-Vlaanderen/groundwater-logger-validation/blob/master/gwloggeR/News.md) file. Each new version should be a top level section, besides the _Development version_ and _Future version_ sections.
* Update the [DESCRIPTION](https://github.com/DOV-Vlaanderen/groundwater-logger-validation/blob/master/gwloggeR/DESCRIPTION) file, mainly the `Version` and `Date` tags.
* Rebuild the [Website] and deploy it on the `gh-pages` branch. (This is the branch that GitHub uses as source for this `gwloggeR` website.)
* Push both `master` and `gh-pages` branch to github.
* Final step is to set the tag for the current version from the shell and push it to github.

```{bash eval=FALSE}
# Note that git-shell can be opened from Rstudio.
git tag x.x.x # make sure to adjust the version.
git push --tags
```
