---
title: "Development guide"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4
    toc: true
    toc_depth: 4
    number_sections: false
    df_print: paged # default, kable, tibble
    fig_caption: true
  pdf_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Development guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Here we describe some  general guidelines for maintenance and development of `gwloggeR`.

## General

`gwloggeR` is developed as part of a larger git-repository: https://github.com/DOV-Vlaanderen/groundwater-logger-validation.

So first we need to checkout this repository somewhere localy.

```{}
git checkout https://github.com/DOV-Vlaanderen/groundwater-logger-validation.git
```

### Projects

The repository consists of 2 project roots:

* [`src/r/groundwater.rproj`](https://github.com/DOV-Vlaanderen/groundwater-logger-validation/blob/master/src/r/groundwater.rproj) which is mainly used for data analysis with R. This project, when opened in Rstudio, will also load [`src/r/.Rprofile`](https://github.com/DOV-Vlaanderen/groundwater-logger-validation/blob/master/src/r/.Rprofile) which will setup the environment for data analysis. Note that here one can also find test-cases for specific versions of `gwloggeR`. For example, [`src/r/outliers/outliers_v0.06_inbo`](https://github.com/DOV-Vlaanderen/groundwater-logger-validation/tree/master/src/r/outliers/outliers_v0.06_inbo) contains all the results based on v0.06 (cf. [#35](https://github.com/DOV-Vlaanderen/groundwater-logger-validation/issues/35)) of the algorithm, which can be used as regression test for new versions of `gwloggeR`.
* [`gwloggeR/gwloggeR.rproj`](https://github.com/DOV-Vlaanderen/groundwater-logger-validation/blob/master/gwloggeR/gwloggeR.Rproj) used for `gwloggeR` package development and testing.

Note that cross-linking of files between projects based on relative paths is allowed since all projects reside in the same repository.

## Documentation

The documentation is built using `roxygen2`. If you are using RStudio and `roxygen2` is installed, then pressing `Ctrl` + `Shift` + `B` will generate:

* Update Rd files (i.e. R documentation files)
* Update NAMESPACE file
* Update vignettes

This command will also install `gwloggeR` in the R `lib\` directory and restart R.

On each update of `gwloggeR` source files, make sure to run this before commiting.

### README.Rmd

This is a special case that must be knitted to `README.md`. In RStudio one can use `Ctrl` + `Shift` + `K`. Knitting must be done before the website is built.

## Website

For `gwloggeR` [website](https://dov-vlaanderen.github.io/groundwater-logger-validation/gwloggeR) we use `pkggdown`. 

We do not use [`travis`](https://pkgdown.r-lib.org/reference/deploy_site_github.html) for automatic website deployment. Each time one updates files that have an impact on the website, one should manualy execute the following commands:

```{r eval=FALSE}
pkgdown::clean_site()
pkgdown::build_site()
```

On each update of `gwloggeR` source files, make sure to run these commands before commiting.
