---
title: "Development guide"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4
    toc: true
    toc_depth: 4
    number_sections: false
    df_print: paged # default, kable, tibble
    fig_caption: true
  pdf_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Development}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Here we describe some  general guidelines for maintenance and development of `gwloggeR`.

## General

`gwloggeR` is developed as part of a larger git-repository: https://github.com/DOV-Vlaanderen/groundwater-logger-validation.

So first we need to checkout this repository somewhere localy.

```{}
git checkout https://github.com/DOV-Vlaanderen/groundwater-logger-validation.git
```

## Documentation

The documentation is built using `roxygen2`. If you are using RStudio and `roxygen2` is installed, then pressing `Ctrl` + `Shift` + `B` will generate:

* Update Rd files (i.e. R documentation files)
* Update NAMESPACE file
* Update vignettes

This command will also install `gwloggeR` in the R `lib\` directory and restart R.

On each update of `gwloggeR` source files, make sure to run this before commiting.

## Website

For `gwloggeR` [website](https://dov-vlaanderen.github.io/groundwater-logger-validation/gwloggeR) we use `pkggdown`. 

We do not use [`travis`](https://pkgdown.r-lib.org/reference/deploy_site_github.html) for automatic website deployment. Each time one updates files that have an impact on the website, one should manualy execute the following commands:

```{r eval=FALSE}
pkgdown::clean_site()
pkgdown::build_site()
```

